---
title: "LaTeX Formulas and Mathematical Notation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LaTeX Formulas and Mathematical Notation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(birtir)
```

## Overview

`birtir` provides `formula_to_latex()` to convert R model formulas into LaTeX mathematical equations. This is useful for model planning, manuscript preparation, and rendering equations directly inside Markdown reports with `md_text()`.

This vignette covers:

- `formula_to_latex()` syntax and options
- Fixed effects and interactions
- Random-effect and mixed-model terms (`(1 | group)`, `(1 + x | group)`, `(x || group)`)
- GLM link functions (`gaussian`, `binomial`, `poisson`, `gamma`)
- Matrix vs expanded equation notation
- Integrating LaTeX math inside `md_text()`

---

## Basic Formula Conversion

`formula_to_latex()` takes a standard R formula and generates a LaTeX equation string:

```{r}
# Intercept-only model
formula_to_latex(y ~ 1)

# Multiple predictors
formula_to_latex(y ~ x1 + x2)
```

### Options

Control subscripting and residual error terms:

```{r}
# Without observation subscripts
formula_to_latex(y ~ x, subscript = FALSE)

# Without explicit error term
formula_to_latex(y ~ x, error_term = FALSE)
```

---

## Interactions and Transformed Terms

Interactions (`x:g` or `x * g`) are converted into LaTeX multiplication terms:

```{r}
# Interaction term
formula_to_latex(y ~ x * g)
```

---

## Mixed-Effects and Random-Effect Formulas

`formula_to_latex()` supports `lme4` and `brms` syntax for random effects:

```{r}
# Random intercept
formula_to_latex(y ~ x + (1 | group))

# Correlated random slope and intercept
formula_to_latex(y ~ x + (1 + x | group))

# Uncorrelated random slope and intercept (||)
formula_to_latex(y ~ x + (x || group))

# Nested grouping factors
formula_to_latex(y ~ x + (1 | g1 / g2))

# Crossed random effects
formula_to_latex(y ~ x + (1 | subject) + (1 | item))
```

---

## Confirmatory Factor Analysis (CFA) and SEM Notation

`birtir` supports `lavaan`-style model syntax with `lavaan_to_latex()` to generate CFA indicator equations, factor loadings, and covariance notation:

```{r}
# Single latent factor (measured by y1, y2, y3)
lavaan_to_latex("F1 =~ y1 + y2 + y3")

# Multifactor CFA with factor covariance
cfa_model <- "
  Visual =~ x1 + x2 + x3
  Text   =~ x4 + x5 + x6
  Visual ~~ Text
"

lavaan_to_latex(cfa_model)

# CFA measurement model in matrix notation
lavaan_to_latex(cfa_model, notation = "matrix")
```

---


## GLMs and Link Functions

Specify model family or explicit link functions to generate link-function equations:

```{r}
# Binomial logistic regression
formula_to_latex(y ~ x, family = "binomial")

# Poisson log-linear regression
formula_to_latex(y ~ x, family = "poisson")

# Gamma regression with inverse link
formula_to_latex(y ~ x, family = "gamma", link = "inverse")
```

---

## Matrix Notation

Use `notation = "matrix"` for model-matrix representation:

```{r}
# Standard linear model in matrix notation
formula_to_latex(y ~ x1 + x2, notation = "matrix")

# Mixed model in matrix notation
formula_to_latex(y ~ x + (1 | group), notation = "matrix")

# Generalized linear mixed model in matrix notation
formula_to_latex(y ~ x + (1 | group), family = "binomial", notation = "matrix")
```

---

## Integrating LaTeX Math in `md_text()`

`md_text()` formats inline text with `glue` interpolation and outputs LaTeX math directly into Markdown reports.

### Interpolating `formula_to_latex()`

```{r}
md_text("Linear model: {formula_to_latex(mpg ~ wt + hp)}")
md_text("Hierarchical model: {formula_to_latex(mpg ~ wt + (1 | cyl))}")
```

### Direct Inline and Display Math

You can write direct LaTeX math in `md_text()`:

```{r}
# Inline math ($...$)
md_text(r"(Estimated coefficient: $\hat{{\beta}}_1 = 2.45$, $p < .001$)")

# Display math ($$...$$)
md_text(r"($$\operatorname{{logit}}(p_i) = \beta_0 + \beta_1 x_i$$)")
```

*Note on Braces:* Because `md_text()` uses `glue::glue()` to evaluate `{...}` expressions, literal curly braces inside raw LaTeX string literals (like `\beta_{0}`) should be doubled (`\beta_{{0}}`) unless generated by `formula_to_latex()`.
