---
title: "Rendering Analysis Scripts and Exporting Reports"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rendering Analysis Scripts and Exporting Reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(birtir)
library(ggplot2)
```

## Overview

`birtir` provides a plain-Markdown render engine for R analysis scripts. Unlike Quarto or R Markdown, `birtir` evaluates standard `.R` scripts line-by-line, captures visible output, saves tables and plots, and writes a clean `.md` report without requiring HTML rendering tools.

This vignette covers the main rendering workflow:

- `render_analysis_md()`
- `md_table()`
- `md_plot()`
- `md_text()`
- `report_labels()`
- `convert_md()`

---

## Writing an Analysis Script

An analysis script is a plain `.R` file that contains standard R code along with `birtir` helper calls:

```r
# scripts/analysis_example.R

birtir::md_text("# Exploratory Analysis\n\nAutomated analysis for mtcars dataset.")

# Standard output is captured in order
summary(mtcars[, c("mpg", "wt", "hp")])

# Format data frames into markdown tables
tbl <- data.frame(
  Metric = c("Mean MPG", "Mean Weight", "Mean HP"),
  Value = c(mean(mtcars$mpg), mean(mtcars$wt), mean(mtcars$hp))
)
birtir::md_table(tbl, caption = "Summary Statistics", digits = 2, note = "Data source: mtcars.")

# Save and embed ggplot objects
p <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()

birtir::md_plot(p, caption = "Miles per Gallon vs Weight", width = 6, height = 4)
```

---

## Rendering to Markdown

Render the script with `render_analysis_md()`:

```r
birtir::render_analysis_md("scripts/analysis_example.R")
```

### Standalone Report Layout (Default)

By default (`layout = "report"`), `render_analysis_md()` creates a dedicated output directory:

```text
outputs/
  analysis-example/
    analysis-example.md
    images/
      analysis-example_fig-001.png
    tables/
      analysis-example_tbl-001.md
```

### Shared Assets ("Book") Layout

When rendering multiple reports into a shared folder, use `layout = "book"`:

```r
birtir::render_analysis_md("scripts/analysis_example.R", layout = "book")
```

This places assets into shared `images/` and `tables/` directories under `outputs/`.

---

## Parametrized Reports and Batch Execution

You can pass named parameter lists to analysis scripts via `params`:

```r
birtir::render_analysis_md(
  script = "scripts/item_template.R",
  report_name = "item-01",
  params = list(item_id = "ITEM_A1", cutoff = 0.70)
)
```

Inside `item_template.R`, parameters are available directly by name (`item_id`, `cutoff`) or as elements of `params`:

```r
birtir::md_text("# Analysis for {item_id}")
birtir::md_text("Filtering with cutoff = {cutoff}")
```

---

## Customizing Captions and Internationalization

Use `report_labels()` to set custom prefixes for table and figure captions:

```r
labels_is <- birtir::report_labels(table = "Tafla", figure = "Mynd")

birtir::render_analysis_md(
  script = "scripts/analysis_example.R",
  labels = labels_is
)
```

---

## Converting Reports to PDF, HTML, or Word

Use `convert_md()` to convert a generated `.md` file into another format via Pandoc:

```r
md_path <- birtir::render_analysis_md("scripts/analysis_example.R")

# Convert to DOCX
birtir::convert_md(md_path, to = "docx")

# Convert to HTML
birtir::convert_md(md_path, to = "html")

# Convert to PDF
birtir::convert_md(md_path, to = "pdf")
```

`convert_md()` automatically strips export-only table links while preserving relative paths to image assets.
