---
title: "Statistical Descriptions and Inline Visualizations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Descriptions and Inline Visualizations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(birtir)
```

## Overview

`birtir` provides formula-driven statistical description tools to summarize variables, group distributions, bivariate relationships, and random-intercept groupings prior to or during report generation.

This vignette covers:

- `describe_data()`
- `as_report_text()`
- `as_report_table()`
- Compact inline text graphics: `inline_hist()`, `inline_boxplot()`, `inline_bar()`, `inline_forest()`, `inline_scatter()`
- Formatting helpers: `fmt_num()`, `fmt_p()`

---

## Formula-based Descriptions with `describe_data()`

### 1. Univariate Description (`x ~ 1`)

Summarize a single continuous or categorical variable:

```{r}
df <- data.frame(
  score = c(41, 53, 49, 52, NA, 47, 60, 55, 38, 44),
  group = factor(c("A", "A", "A", "B", "B", "B", "A", "B", "A", "B"))
)

# Continuous univariate summary
desc_uni <- describe_data(df, score ~ 1)
print(desc_uni)
```

### 2. Grouped Continuous Summary (`y ~ group`)

Summarize continuous distributions across categorical groups:

```{r}
desc_grp <- describe_data(df, score ~ group)
print(desc_grp)
```

### 3. Additive and Interaction Terms (`y ~ g + h` and `y ~ g:h`)

Additive terms provide separate group summaries; interaction terms compute crossed grouping combinations:

```{r}
df2 <- data.frame(
  score = c(12, 15, 18, 22, 25, 30, 28, 35),
  group = factor(c("A", "A", "A", "A", "B", "B", "B", "B")),
  condition = factor(c("Low", "Low", "High", "High", "Low", "Low", "High", "High"))
)

# Additive grouped summary
describe_data(df2, score ~ group + condition)

# Crossed interaction summary
describe_data(df2, score ~ group:condition)
```

### 4. Continuous Bivariate Relationships (`y ~ x`)

Compute correlations, regression slopes, and 95% confidence intervals:

```{r}
df_rel <- data.frame(
  x = c(1, 2, 3, 4, 5, 6),
  y = c(10, 14, 19, 21, 28, 32)
)

desc_rel <- describe_data(df_rel, y ~ x)
print(desc_rel)
```

### 5. Random-Intercept Groupings (`y ~ (1 | group)`)

Compact forest-style summaries for hierarchical or grouped grouping factors:

```{r}
describe_data(df, score ~ (1 | group))
```

---

## Extracting Report Text and Report Tables

Convert any `birtir_description` object directly into report-ready text or structured tibbles:

```{r}
# Convert to formatted text with APA-style italicized statistics
as_report_text(desc_uni)

# Convert to a tidy report table
as_report_table(desc_grp)
```

Inside `render_analysis_md()`, you can pass these directly to `md_text()` and `md_table()`:

```r
desc <- describe_data(df, score ~ group)
md_text(as_report_text(desc))
md_table(as_report_table(desc), caption = "Score by Group")
```

---

## Text-Based Inline Graphics

`birtir` includes zero-dependency, character-rendered plot helpers for console previews and Markdown output.

### Inline Histogram (`inline_hist`)

```{r}
inline_hist(df$score[!is.na(df$score)], n_bins = 5)
```

### Inline Boxplot (`inline_boxplot`)

```{r}
inline_boxplot(df$score[!is.na(df$score)], width = 30)
```

### Inline Bar Chart (`inline_bar`)

```{r}
inline_bar(table(df$group), width = 15, digits = 0)
```

### Inline Scatter Plot (`inline_scatter`)

```{r}
inline_scatter(df_rel$x, df_rel$y, width = 20, height = 5)
```

### Inline Forest Plot (`inline_forest`)

```{r}
inline_forest(
  estimate = c(x1 = 0.31, x2 = -0.15),
  ci_low = c(0.12, -0.35),
  ci_high = c(0.50, 0.05),
  n = c(100, 100),
  width = 12
)
```

---

## Formatting Numbers and p-Values

Format numbers for reporting with APA 7 guidelines:

```{r}
# APA-style numbers (optional leading zero suppression)
fmt_num(c(0.123, -0.456, 12.0), digits = 2)
fmt_num(0.123, digits = 2, drop_leading_zero = TRUE)

# APA-style p-values
fmt_p(c(0.042, 0.0003, NA))
```
